---
title: "MOSClip Two-Class Analysis on TCGA ovarian cancer patients"
output: html_document
---

## Prepare data for MOSClip analysis

Now we are almost ready to run a two-class analysis with MOSClip!
First of all, we can load necessary libraries and the pre-processed ovarian cancer data 
that we downloaded in the previous tutorials from TCGA.
We will also set a seed, in order to have reproducible results in case of future repetition of the analysis.

```{r message=FALSE}
library(org.Hs.eg.db)
library(EDASeq)
library(graphite)
library(MOSClip)
library(kableExtra)


load("downloadTCGA/TCGA-OV-pre-processed.RData")

set.seed(1234)
```

We need to prepare data in order to run MOSClip. 
The first step is to modify all the multi-omics matrices assigning the type of gene identifier. 
Since we will use `graphite` to download a list of pathways and their graphical structure, 
we need to format gene names for each omic in order to be compatible with `graphite` gene names. 
Here, we will work with Entrez Gene ID, thus we have to indicate with the prefix tag “ENTREZID:" each gene.

```{r}
expression <- expAvg
row.names(expression) <- paste0("ENTREZID:", row.names(expression))
mutation <- ov.mutations$data
row.names(mutation) <- paste0("ENTREZID:", row.names(mutation))
names(metClustValues$eMap) <- paste0("ENTREZID:", row.names(metClustValues$eMap))
row.names(ov.cnv) <- paste0("ENTREZID:", row.names(ov.cnv))
```

Moving to patient selection, we will keep only patients whose class annotation 
is available and their intersection with available patients across the 4 omics.
Finally, our class annotation dataframe will be filtered to keep only selected patients.

```{r}
# select common patients
patients <- row.names(classes)
patients <- intersect(patients, colnames(expression))
patients <- intersect(patients, colnames(metClustValues$MET_Cancer_Clustered))
patients <- intersect(patients, colnames(mutation))
patients <- intersect(patients, colnames(ov.cnv))

classAnnot <- classes[patients, , drop=FALSE]

table(classAnnot)
```

At this point, we need to extract selected patients for each multi-omic matrix. 
After patient selection, we can normalize (upper quartile normalization from `EDASeq` package) 
and log-transform expression data.

```{r}
# normalize expression data
expression <- expression[, patients, drop = F]
keep = apply(expression >= 100, 1, any)
expNorm <- betweenLaneNormalization(expression[keep, , drop = F], which = "upper")
pseudoExpNorm <- log2(expNorm + 1)

methylation <- metClustValues
methylation$MET_Cancer_Clustered <- methylation$MET_Cancer_Clustered[, patients, 
    drop = F]

mutation <- mutation[, patients, drop = F]
cnv <- ov.cnv[, patients, drop = F]
```

We are now ready to generate an object of class `Omics` using MOSClip function `makeOmics`.
This object is necessary to run each type of `MOSClip` analyses. 
It is based on `MultiAssayExperiment` object, containing an `ExperimentList` with matrices for each omic 
(we suggest to use standard names for each experiment as shown in this example).
`colData` will contain class annotation for patients. Additionally, specific slots 
for `MOSClip` analysis exist, including `modelInfo`, where the user must specify 
the desired method for data reduction for each omic, and `specificArgs`, with specific
parameters to be used by reduction functions. 
Both these slots must have the same dimension as `ExperimentList`.

The list of available methods for data dimensionality reduction is easily made available 
with `availableOmicsMethods()`.

```{r}
multiOmics <- makeOmics(experiments = list(exp = pseudoExpNorm, 
                                           met = methylation$MET_Cancer_Clustered, 
                                           mut = mutation, 
                                           cnv = as.matrix(cnv)), 
                        colData = classAnnot,
    modelInfo = c("summarizeWithPca", "summarizeInCluster", 
    "summarizeToNumberOfEvents", "summarizeToNumberOfDirectionalEvents"), 
    specificArgs = list(pcaArgs = list(name = "exp", shrink = "FALSE", method = "sparse", maxPCs = 3),
                        clusterArgs = list(name = "met", dict = methylation$eMap, max_cluster_number = 3), 
                        countEvent = list(name = "mut", min_prop = 0.05), 
                        cnvAgv = list(name = "cnv", min_prop = 0.05)))
```

### Download Reactome pathways

To run a `MOSClip` analysis we also need a list of pathways that we want to test. 
We decided to use pathways collected in Reactome database. They can be downloaded 
using `graphite` that also is able to convert gene identifiers (here we are using 
Entrez Gene ID). This process may take many minutes, for this reason we will save
the final `PathwayList` object in a specific directory, ready to be used in future analyses.

To avoid redundant and too heavy analyses, we select Reactome pathways based on their 
number of nodes, considering only nodes that are present at least in the expression matrix.
We prepare two distinct `PathwayList` objects: 
* one containing pathways with more than 20 and less than 100 nodes,  
* a larger one with pathways having more than 10 and less than 700 nodes. 

```{r}
if (file.exists("downloadTCGA/reactome-entrez-2024-05-27.RData")) {
  load("downloadTCGA/reactome-entrez-2024-05-27.RData")
} else {
  reactome <- pathways("hsapiens", "reactome")
  reactome <- convertIdentifiers(reactome, "entrez")
  file = paste0("downloadTCGA", "/reactome-entrez-", as.character(Sys.Date()), ".RData")
  save(reactome, file = file)
}

nodesLength <- sapply(reactome, function(g) {length(intersect(graphite::nodes(g), row.names(pseudoExpNorm)))})
reactSmall <- reactome[nodesLength >= 20 & nodesLength <= 100]
reactHuge <- reactome[nodesLength >= 10 & nodesLength <= 700]
```

### Prepare class annotations

We define the classes that we want to compare. Here we are using the subtypes 
defined for TCGA ovarian cancer patients. We think it could be interesting to compare
on a multi-omic level immunoreactive patients with mesenchymal patients. The user 
can decide which subtypes to compare.

We can filter accordingly our class annotation data frame and `multiOmics` object.

```{r}
class1 <- "Immunoreactive"
class2 <- "Mesenchymal"
classAnnotation <- classAnnot[classAnnot$classes %in% c(class1, class2), , drop=FALSE]
multiOmics <- multiOmics[,row.names(classAnnotation)]
```

Now we are ready for MOSClip two-class analysis.

## Two-class analysis on modules

```{r message=FALSE, warning=FALSE}
if (file.exists("MOSresults/twoClassM.rds")){
  twoClassM <- readRDS("MOSresults/twoClassM.rds")
} else { 
    twoClassM <- lapply(reactSmall, function(g) {
        res <- multiOmicsTwoClassModuleTest(multiOmics, g, classAnnotation, useTheseGenes = row.names(pseudoExpNorm))
        res
    })
    saveRDS(twoClassM, file="MOSresults/twoClassM.rds")
}

moduleSummary <- multiPathwayModuleReport(twoClassM)
```

```{r echo=FALSE}
kable(moduleSummary[1:100,-1]) %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "600px")
```


### Plots

```{r}
plotModuleReport(twoClassM[["Activation of Matrix Metalloproteinases"]])
```

```{r message=FALSE}
plotModuleInGraph(twoClassM[["Activation of Matrix Metalloproteinases"]], reactSmall, 2)
```


```{r message=FALSE}
plotModuleHeat(twoClassM[["Activation of Matrix Metalloproteinases"]], 2, additionalAnnotations = classAnnotation, 
               additionalPaletteNames = list(classes="violet"))
```


### Permutations

```{r echo=FALSE}
load("../downloadTCGA/twoClassSub_immmes.RData")
moduleSummary <- multiPathwayModuleReport(twoClassM)
```


```{r warning=FALSE}
useThisPathways <- unique(moduleSummary$pathway[moduleSummary$pvalue <= 0.05])
sModule <- moduleSummary[moduleSummary$pathway %in% useThisPathways, , drop = T]

if (file.exists("MOSresults/permsM.RData")){
    load("MOSresults/permsM.RData")
}else{
    perms <- resamplingModulesTwoClass(fullMultiOmics = multiOmics, 
                                               classAnnotation, reactSmall, nperm = 2, 
                                               pathwaySubset = useThisPathways, 
                                               genesToConsider = row.names(pseudoExpNorm))
    save(perms, file="MOSresults/permsM.RData")
}

stableModulesSummary <- selectStablePathwaysModules(perms = perms, moduleSummary = sModule, success = 80)
resamplingSuccessCount <- getPathwaysModulesSuccess(perms = perms, moduleSummary = sModule)
moduleSummary <- addResamplingCounts(moduleSummary, resamplingSuccessCount)
```

### More plots

```{r}
runSupertest(moduleSummary, pvalueThr = 0.05, zscoreThr = 0.05, resampligThr = 80,
             excludeColumns = c("pathway", "module", "resamplingCount"))
```

```{r}
pathHierarchyGraph <-  igraph::graph_from_data_frame(d =downloadPathwayRelationFromReactome(), directed = TRUE)

omicsClasses2pathways <- computeOmicsIntersections(moduleSummary, pvalueThr = 0.05, 
                                                   zscoreThr = 0.05, resampligThr = 80, 
                                                   excludeColumns = c("pathway", "module", "resamplingCount"))
omicsClasses2pathways <- lapply(omicsClasses2pathways, stripModulesFromPathways)
omicsClasses2fathers <- lapply(omicsClasses2pathways, annotePathwayToFather, graphiteDB = reactome, hierarchy = pathHierarchyGraph)
correspondence <- lapply(names(omicsClasses2pathways), function(omicClass) {
  data.frame(path = omicsClasses2pathways[[omicClass]], father = omicsClasses2fathers[[omicClass]], 
             stringsAsFactors = F)})
MOMfreqDataframe <- computeFreqs(omicsClasses2fathers)
combiClass <- grep(";", MOMfreqDataframe$class)
MOMfreqDataframe.multi <- MOMfreqDataframe[combiClass, , drop = F]
colors <- c(`cnv;exp` = "#ff5722", `cnv;exp;met` = "#FFAB00", `cnv;met` = "#A5F037", 
            `cnv;mut` = "#30B30D", `exp;met` = "#6c5b07", `exp;mut` = "#A200AD", `met;mut` = "#01B7BF")

plotFrequencies(MOMfreqDataframe.multi, minSize = 6, maxSize = 8, width = 10, manualColors = colors, lineSize = 1)
```

## Two-class analysis on pathways

```{r warning=FALSE}
multiOmics@specificArgs$pcaArgs$shrink=TRUE
multiOmics@specificArgs$pcaArgs$method="topological"

if (file.exists("MOSresults/twoClassP.rds")){
  twoClassP <- readRDS("MOSresults/twoClassP.rds")
} else {
  twoClassP <- lapply(reactHuge, function(g) {
    res <- multiOmicsTwoClassPathwayTest(multiOmics, g, classAnnotation, 
                                         useTheseGenes = row.names(pseudoExpNorm))
    res
  })
  saveRDS(twoClassP, file="MOSresults/twoClassP.rds")
 }

dup <- which(names(twoClassP)=="Translation of Replicase and Assembly of the Replication Transcription Complex")
names(twoClassP)[dup[2]] <- "Translation of Replicase and Assembly of the Replication Transcription Complex.1"


pathwaySummary <- multiPathwayReport(twoClassP)
```

```{r}
kable(pathwaySummary[1:100,-1]) %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "600px")
```


```{r message=FALSE}
plotPathwayHeat(twoClassP[["Activation of Matrix Metalloproteinases"]], additionalAnnotations = classAnnotation,
                additionalPaletteNames = list(classes="violet"))
```

```{r}
runSupertest(pathwaySummary, pvalueThr = 0.05, zscoreThr = 0.05)
```

```{r}
omicsClasses2pathways <- computeOmicsIntersections(pathwaySummary)
omicsClasses2fathers <- lapply(omicsClasses2pathways, annotePathwayToFather, graphiteDB = reactome, 
    hierarchy = pathHierarchyGraph)
freqDataframe <- computeFreqs(omicsClasses2fathers)

colors = c(cnv = "#E5C002", `cnv;exp` = "#ff5722", `cnv;exp;mut` = "#6B213A", `cnv;met` = "#A5F037", 
    exp = "#E33402", `exp;met` = "#6c5b07", `exp;mut` = "#A200AD", met = "#04B880", 
    mut = "#3871CC")

plotFrequencies(freqDataframe, minSize = 3, maxSize = 7, width = 9, manualColors = colors, 
    lineSize = 1)
```


### Permutations

```{r warning=FALSE, eval=FALSE}
useThisPathways <- unique(row.names(pathwaySummary)[pathwaySummary$pvalue <= 0.05])
sPathway <- pathwaySummary[row.names(pathwaySummary) %in% useThisPathways, , drop = T]

if (file.exists("MOSresults/permsP.RData")){
    load("MOSresults/permsP.RData")
} else{
    perms <- resamplingPathwayTwoClass(fullMultiOmics = multiOmics, 
                                                classAnnotation, reactHuge, nperm = 100, 
                                                pathwaySubset = useThisPathways, 
                                                genesToConsider = row.names(pseudoExpNorm))
    save(perms, file="MOSresults/permsP.RData")
}


stablePathwaysSummary <- selectStablePathwaysModules(perms = perms, moduleSummary = sPathway, success = 80)
resamplingSuccessCount <- getPathwaysModulesSuccess(perms = perms, moduleSummary = sPathway)
pathwaySummary <- addResamplingCounts(pathwaySummary, resamplingSuccessCount)
```